### 2023-01-24 manuscript.tex
peerj citation style [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/86bc7fe899c7464fad8f4ebe872efb61c3be1394.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/86bc7fe899c7464fad8f4ebe872efb61c3be1394/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4000951091

This manuscript
(\href{https://stain.github.io/2022-fdo-paper/v/51048f35d5fce97aa4194e6426fdd7c704f19199/}{permalink})
was automatically generated
from \href{https://github.com/stain/2022-fdo-paper/tree/51048f35d5fce97aa4194e6426fdd7c704f19199}{stain/2022-fdo-paper@51048f3}
on January 24, 2023.

@@

The FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures. They have been widely recognised by the research community and policy makers as a goal to strive for. In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures {[}\protect\hyperlink{ref-u8Cy0psL}{2}{]}. The EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.

Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures. The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.

In other discourse, Linked Data {[}\protect\hyperlink{ref-19s3yyxDn}{5}{]} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles {[}\protect\hyperlink{ref-GJHYN6xW}{6},\protect\hyperlink{ref-EJDjHrUv}{7}{]}. Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully {[}\protect\hyperlink{ref-jHwttlTs}{8}{]}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services {[}\protect\hyperlink{ref-99MS5xE0}{9},\protect\hyperlink{ref-J2H8yssV}{10}{]}. While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability {[}\protect\hyperlink{ref-DNTCUjpC}{11},\protect\hyperlink{ref-Pa67pUtR}{12}{]}.

One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality {[}\protect\hyperlink{ref-uxWjmlio}{13},\protect\hyperlink{ref-v0WwFHgT}{14}{]} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves {[}\protect\hyperlink{ref-5V3GqSzR}{15}{]}.

@@

The FAIR Digital Object Forum {[}\protect\hyperlink{ref-DMEBhHID}{16}{]} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public. As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:

\textbf{FAIR Digital Object Overview and Specifications} {[}\protect\hyperlink{ref-11NpETLq1}{17}{]} is a comprehensive overview of FAIR Digital Object specifications listed below. It serves as a primer that introduces FDO concepts and the remaining documents.

The \textbf{FDO Forum Document Standards} {[}\protect\hyperlink{ref-TQdku4YF}{18}{]} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions. In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.

The \textbf{FDO Requirement Specifications} {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} is an update of {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} as the foundational definition of FDO. This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations. The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.

The \textbf{Machine actionability} {[}\protect\hyperlink{ref-iJeys0T5}{21}{]} sets out to define what is meant by \emph{machine actionability} for FDOs. \emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar. The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.

\textbf{Configuration Types} {[}\protect\hyperlink{ref-SPd0uEqO}{22}{]} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs. Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]}

\textbf{PID Profiles \& Attributes} {[}\protect\hyperlink{ref-ljVViWCl}{24}{]} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile. This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}. This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.

\textbf{Kernel Attributes \& Metadata} {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc. This document expands on RDA Recommendation on PID Kernel Information {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}. It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.

\textbf{Granularity, Versioning, Mutability} {[}\protect\hyperlink{ref-16W9AycSW}{27}{]} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements. The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.

\textbf{DOIP Endorsement Request} {[}\protect\hyperlink{ref-15yGpJ0wh}{28}{]} is an endorsement of the DOIP v2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} specification as a potential FDO implementation, as it has been applied by several institutions {[}\protect\hyperlink{ref-jRWurmQc}{30}{]}. The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).

\textbf{Upload of FDO} {[}\protect\hyperlink{ref-1H7cmmr69}{31}{]} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve). ResourceSync {[}\protect\hyperlink{ref-eS5y9TRh}{32}{]} is suggested as one type of service to list FDOs. This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).

\textbf{Typing FAIR Digital Objects} {[}\protect\hyperlink{ref-ZFzPxCNB}{33}{]} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose. This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types. Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work. How to define such FDO operations is not specified.

\textbf{Implementation of Attributes, Types, Profiles and Registries} {[}\protect\hyperlink{ref-126uxr5pI}{34}{]} details how to establish FDO registries for types and FDO profiles, with their association with PID systems. This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices. Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} {[}\protect\hyperlink{ref-1CqIZ47pu}{35}{]}.

@@

The concept of \textbf{FAIR Digital Objects} {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} has been introduced as way to expose research data as active objects that conform to the FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]}. This builds on the \emph{Digital Object} (DO) concept {[}\protect\hyperlink{ref-11MnuwJ4l}{36}{]}, first introduced in 1995 {[}\protect\hyperlink{ref-3Uqe3fuK}{37}{]} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles. DO was the inspiration for the ITU X.1255 framework {[}\protect\hyperlink{ref-103Hw8H43}{38}{]} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 {[}\protect\hyperlink{ref-16uB3jxpa}{39}{]}.

In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}. The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs). The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing {[}\protect\hyperlink{ref-tz0P3DTC}{40}{]}.

Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}. Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}. The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).

@@

FDO is an evolving concept. A set of FDO Demonstrators {[}\protect\hyperlink{ref-jRWurmQc}{30}{]} highlight how current adapters are approaching implementations of FDO from different angles:

\begin{itemize}
\tightlist
\item
  Building on the Digital Object concept, using the simplified DOIP v2 specification {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, see {[}\protect\hyperlink{ref-xuQKRRx5}{41}{]}.} (usually TCP/IP over TLS). The main DOIP operations are retrieving, creating and updating digital objects. These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}. FDO types are registered in the local Cordra instance, where they are specified using JSON Schema {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]} and PIDs are assigned using the Handle system. Several type registries have been established.
\item
  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
\item
  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
\end{itemize}

@@

The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee {[}\protect\hyperlink{ref-Guy7cjVY}{43}{]}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs. The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}. The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.

Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) {[}\protect\hyperlink{ref-16WGmCih8}{44}{]} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format {[}\protect\hyperlink{ref-16wvFH4nh}{45}{]} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).

While triple-based knowledge representations were not new {[}\protect\hyperlink{ref-NxvhwSmG}{46}{]}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs {[}\protect\hyperlink{ref-RRtwPF7Y}{47}{]} are generalised forms of URLs that include locator-less identifiers
  such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years {[}\protect\hyperlink{ref-nW4HY8Nq}{48}{]}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) {[}\protect\hyperlink{ref-HqZ3r6J3}{49}{]}. RDF 1.1 formats use Unicode to support IRIs {[}\protect\hyperlink{ref-rqjuXkEf}{50}{]}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} {[}\protect\hyperlink{ref-rnzJ4khD}{51}{]}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.

@@

The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is). More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} {[}\protect\hyperlink{ref-yd2WE9eq}{52}{]} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.

Established Semantic Web services and standards include SPARQL {[}\protect\hyperlink{ref-xww8fb1B}{53}{]} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).

>>>>

This manuscript
(\href{https://stain.github.io/2022-fdo-paper/v/86bc7fe899c7464fad8f4ebe872efb61c3be1394/}{permalink})
was automatically generated
from \href{https://github.com/stain/2022-fdo-paper/tree/86bc7fe899c7464fad8f4ebe872efb61c3be1394}{stain/2022-fdo-paper@86bc7fe}
on January 24, 2023.

@@

The FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}) encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures. They have been widely recognised by the research community and policy makers as a goal to strive for. In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures (\protect\hyperlink{ref-u8Cy0psL}{Mons et al., 2017}). The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.

Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures. The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.

In other discourse, Linked Data (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\protect\hyperlink{ref-EJDjHrUv}{Bonino Da Silva Santos et al., 2016}; \protect\hyperlink{ref-GJHYN6xW}{Hasnain \& Rebholz-Schuhmann, 2018}). Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}), opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\protect\hyperlink{ref-99MS5xE0}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021a}; \protect\hyperlink{ref-J2H8yssV}{Neumann, Laranjeiro \& Bernardino, 2021}). While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022}).

One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\protect\hyperlink{ref-uxWjmlio}{Stefi, 2015}; \protect\hyperlink{ref-v0WwFHgT}{Stefi \& Hess, 2015}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves (\protect\hyperlink{ref-5V3GqSzR}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021b}).

@@

The FAIR Digital Object Forum (\protect\hyperlink{ref-DMEBhHID}{{`FAIR Digital Objects Forum \textbar{}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public. As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:

\textbf{FAIR Digital Object Overview and Specifications} (\protect\hyperlink{ref-11NpETLq1}{Anders et al., 2022a}) is a comprehensive overview of FAIR Digital Object specifications listed below. It serves as a primer that introduces FDO concepts and the remaining documents.

The \textbf{FDO Forum Document Standards} (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions. In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.

The \textbf{FDO Requirement Specifications} (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) is an update of (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) as the foundational definition of FDO. This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations. The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.

The \textbf{Machine actionability} (\protect\hyperlink{ref-iJeys0T5}{Weiland et al., 2022a}) sets out to define what is meant by \emph{machine actionability} for FDOs. \emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar. The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.

\textbf{Configuration Types} (\protect\hyperlink{ref-SPd0uEqO}{Lannom et al., 2022a}) classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs. Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022})

\textbf{PID Profiles \& Attributes} (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}) specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile. This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}). This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.

\textbf{Kernel Attributes \& Metadata} (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc. This document expands on RDA Recommendation on PID Kernel Information (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}). It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.

\textbf{Granularity, Versioning, Mutability} (\protect\hyperlink{ref-16W9AycSW}{Working Group, 2022b}) considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements. The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.

\textbf{DOIP Endorsement Request} (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) is an endorsement of the DOIP v2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) specification as a potential FDO implementation, as it has been applied by several institutions (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}). The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).

\textbf{Upload of FDO} (\protect\hyperlink{ref-1H7cmmr69}{Blanchi et al., 2022a}) illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve). ResourceSync (\protect\hyperlink{ref-eS5y9TRh}{{`ResourceSync Framework Specification - Table of Contents'}}) is suggested as one type of service to list FDOs. This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).

\textbf{Typing FAIR Digital Objects} (\protect\hyperlink{ref-ZFzPxCNB}{Lannom et al., 2022b}) defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose. This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types. Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work. How to define such FDO operations is not specified.

\textbf{Implementation of Attributes, Types, Profiles and Registries} (\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) details how to establish FDO registries for types and FDO profiles, with their association with PID systems. This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices. Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} (\protect\hyperlink{ref-1CqIZ47pu}{Schwardmann \& Kálmán, 2022}).

@@

The concept of \textbf{FAIR Digital Objects} (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) has been introduced as way to expose research data as active objects that conform to the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}). This builds on the \emph{Digital Object} (DO) concept (\protect\hyperlink{ref-11MnuwJ4l}{Kahn \& Wilensky, 2006}), first introduced in 1995 (\protect\hyperlink{ref-3Uqe3fuK}{Kahn \& Wilensky, 1995}) as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles. DO was the inspiration for the ITU X.1255 framework (\protect\hyperlink{ref-103Hw8H43}{{`X.1255 : Framework for discovery of identity management information'}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}).

In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}. The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs). The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing (\protect\hyperlink{ref-tz0P3DTC}{Wittenburg et al., 2019}).

Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}), in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}. Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}). The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).

@@

FDO is an evolving concept. A set of FDO Demonstrators (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) highlight how current adapters are approaching implementations of FDO from different angles:

\begin{itemize}
\tightlist
\item
  Building on the Digital Object concept, using the simplified DOIP v2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), see (\protect\hyperlink{ref-xuQKRRx5}{{`DOIP and Examples --- Cordra documentation'}}).} (usually TCP/IP over TLS). The main DOIP operations are retrieving, creating and updating digital objects. These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}. FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}}) and PIDs are assigned using the Handle system. Several type registries have been established.
\item
  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
\item
  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
\end{itemize}

@@

The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee (\protect\hyperlink{ref-Guy7cjVY}{Berners-Lee \& Fischetti, 1999}), at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs. The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}. The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.

Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\protect\hyperlink{ref-16WGmCih8}{{`RDF 1.1 Primer'}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\protect\hyperlink{ref-16wvFH4nh}{{`Resource Description Framework (RDF) Model and Syntax Specification'}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).

While triple-based knowledge representations were not new (\protect\hyperlink{ref-NxvhwSmG}{Stanczyk, 1987}), the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs (\protect\hyperlink{ref-RRtwPF7Y}{Berners-Lee, Fielding \& Masinter, 2005}) are generalised forms of URLs that include locator-less identifiers
  such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years (\protect\hyperlink{ref-nW4HY8Nq}{{`"info" URI Registry (Frozen)'}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) (\protect\hyperlink{ref-HqZ3r6J3}{Juty, Le Novere \& Laibe, 2011}). RDF 1.1 formats use Unicode to support IRIs (\protect\hyperlink{ref-rqjuXkEf}{Duerst \& Suignard, 2005}), which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.

@@

The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is). More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} (\protect\hyperlink{ref-yd2WE9eq}{Horrocks \& Hendler, 2002}) gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.

Established Semantic Web services and standards include SPARQL (\protect\hyperlink{ref-xww8fb1B}{{`SPARQL 1.1 Overview'}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
